---
title: "Confidence Intervals"
subtitle: "using mtcars dataset  \n  \nLecture by Christine Mallot - DSTI S21"
author: "Benoit Mialet"
date: "20/09/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# I) Sample and population variance

```{r}
library(prettyR) # for method valid.n()
```
The two following estimators for $\sigma^2$ are used to build confidence intervals (CI).



## I.1) Sample variance
* If we work on a sample of the whole population, **expectation $\mu$ is unknown** and estimated by $\overline{X_n}$. In this case, **sample variance $\hat{\sigma_{n}}^2$** is an unbiased and consistent estimator for $\sigma^2$:

$$\hat{\sigma_{n}}^2 = \frac{1}{n-1}\sum_{i=1}^{n}(X_i-\overline{X_n})^2$$
```{r}
X <- mtcars$wt        # random variable
var(X, na.rm=TRUE)    # sample variance
sd(X, na.rm=TRUE)     # sample standard deviation
```
  



## I.2) Population variance

* If **expectation $\mu$ is known**, **population variance $\sigma^2$** is an unbiased estimator for $\sigma^2$:

$$\tilde{\sigma_{n}}^2 = \frac{1}{n}\sum_{i=1}^{n}(X_i-\mu)^2$$

```{r}
X <- mtcars$wt                 # random variable
mu <- 3.2                      # expectation (arbitrary value)       
pop_var <- 1/(valid.n(X, na.rm=TRUE))*sum((X - mu)^2)    # population variance
pop_sd <- sqrt(pop_var)                                  # population standard deviation
pop_var
pop_sd
```
  
  
  

# I) Confidence interval of the expectation $\mu$

```{r}
library(prettyR) # for method valid.n()
```

* some terms:
  + $X$: random variable
  + $\overline{X_n}$: empirical mean for n individuals 
  + $\sigma^2$: variance
  + $\hat{\sigma}^2$: sample variance (estimator for $\sigma^2$)
  + $\tilde{\sigma}^2$: population variance (estimator for $\sigma^2$)
  + $1-\alpha$: confidence level
  + $Z_{(1-\alpha/2)}$: quantile of order $1-\alpha/2$ for a Normal random variable
  + $t_{(1-\alpha/2; n-)}$: quantile of order $1-\alpha/2$ for a Student random variable with $n-1$ degree of freedom  





## I.1) If $X$ is not assumed to be gaussian, and variance $\sigma^2$ unknown (most frequent case):

**Asymptotic** symmetric CI, thanks to the Central Limit Theorem, is: 



$$\left[ \overline{X_n} - \frac{\hat{\sigma_{n}}}{\sqrt{n}}Z_{(1-\alpha/2)} ; \overline{X_n} + \frac{\hat{\sigma_{n}}}{\sqrt{n}}Z_{(1-\alpha/2)} \right] $$
With $\overline{X_n}$ the empirical mean and $\hat{\sigma_{n}}$ the population standard deviation


```{r}
X <- mtcars$wt        # random variable
alpha <- 0.05         # confidence level = 1-alpha = 0.95
IC_min <- mean(X, na.rm=TRUE)-sd(X, na.rm=TRUE)/sqrt(valid.n(X, na.rm=TRUE))*qnorm(1-alpha/2)
IC_max <- mean(X, na.rm=TRUE)+sd(X, na.rm=TRUE)/sqrt(valid.n(X, na.rm=TRUE))*qnorm(1-alpha/2)
#Asymptotic CI:
mean(X)
IC_min 
IC_max
```
  


## I.2) If X is  assumed to be gaussian, and variance $\sigma^2$ known:


**Asymptotic** symmetric CI is:



$$\left[ \overline{X_n} - \frac{\sigma}{\sqrt{n}}Z_{1-\alpha/2} ; \overline{X_n} + \frac{\sigma}{\sqrt{n}}Z_{1-\alpha/2} \right] $$
With $\overline{X_n}$ the empirical mean  



```{r}
X <- mtcars$wt        # random variable
alpha <- 0.05         # confidence level = 1-alpha = 0.95
sd <- 0.9             # standard deviation (arbitrary value)
IC_min <- mean(X, na.rm=TRUE)-sd/sqrt(valid.n(X, na.rm=TRUE))*qnorm(1-alpha/2)
IC_max <- mean(X, na.rm=TRUE)+sd/sqrt(valid.n(X, na.rm=TRUE))*qnorm(1-alpha/2)
#Asymptotic CI:
mean(X)
IC_min
IC_max
```
  


## I.3) If X **is gaussian**, and **variance** $\sigma^2$ **unknown**:

**Exact** symmetric CI is:
 


$$\left[ \overline{X_n} - \frac{\hat{\sigma_{n}}}{\sqrt{n}}t_{(1-\alpha/2;n-1)} ; \overline{X_n} + \frac{\hat{\sigma_{n}}}{\sqrt{n}}t_{(1-\alpha/2;n-1)} \right] $$
With $\overline{X_n}$ the empirical mean and $\hat{\sigma_{n}}$ the sample standard deviation


```{r}
X <- mtcars$wt        # random variable
alpha <- 0.05         # confidence level = 1-alpha = 0.95
IC_min <- mean(X, na.rm=TRUE)-sd(X, na.rm=TRUE)/sqrt(valid.n(X, na.rm=TRUE))*qt((1-alpha/2), (valid.n(X, na.rm=TRUE)-1))
IC_max <- mean(X, na.rm=TRUE)+sd(X, na.rm=TRUE)/sqrt(valid.n(X, na.rm=TRUE))*qt((1-alpha/2), (valid.n(X, na.rm=TRUE)-1))

#Exact CI:
mean(X)
IC_min
IC_max
```
  



## I.4) If X **is gaussian**, and **variance** $\sigma^2$ **known**:

**Exact** symmetric CI is:
 



$$\left[ \overline{X_n} - \frac{\sigma}{\sqrt{n}}Z_{1-\alpha/2} ; \overline{X_n} + \frac{\sigma}{\sqrt{n}}Z_{1-\alpha/2} \right] $$
With $\overline{X_n}$ the empirical mean


```{r}
X <- mtcars$wt        # random variable
alpha <- 0.05         # confidence level = 1-alpha = 0.95
sd <- 0.9             # standard deviation (arbitrary value)
IC_min <- mean(X, na.rm=TRUE)-sd/sqrt(valid.n(X, na.rm=TRUE))*qnorm((1-alpha/2), (valid.n(X, na.rm=TRUE)-1))
IC_max <- mean(X, na.rm=TRUE)+sd/sqrt(valid.n(X, na.rm=TRUE))*qnorm((1-alpha/2), (valid.n(X, na.rm=TRUE)-1))
#Exact CI:
mean(X)
IC_min
IC_max
```
  


# II)  Confidence interval of the variance $\sigma^2$
## (Gaussian distribution only)


```{r}
library(prettyR) # for method valid.n()
```

## II.1) If expectation $\mu$ is unknown (most often case):

We use **sample variance $\hat{\sigma_{n}}^2$** (see I.1)). Confidence interval is:

$$\left[ \frac{\hat{\sigma_{n}}^2.n}{C_{(1-\alpha_2;n-1)}} ; \frac{\hat{\sigma_{n}}^2.n}{C_{(\alpha_1;n-1)}} \right] $$
With $\alpha_1 + \alpha_2 = \alpha$

```{r}
X <- mtcars$wt        # random variable
alpha <- 0.05         # confidence level = 1-alpha = 0.95
n <- (valid.n(X, na.rm=TRUE))
var(X)                # sample variance

# exact CI
IC_min <- var(X)*n/qchisq((1-alpha/2), n-1)
IC_max <- var(X)*n/qchisq((alpha/2), n-1)
IC_min
IC_max
```



## II.2) If expectation $\mu$ is known:

We use **population variance** $\tilde{\sigma_{n}}^2$ (see I.2)). Confidence interval is:

$$\left[ \frac{\tilde{\sigma_{n}}^2.n}{C_{(1-\alpha_2;n)}} ; \frac{\tilde{\sigma_{n}}^2.n}{C_{(\alpha_1;n)}} \right] $$
With $\alpha_1 + \alpha_2 = \alpha$

```{r}
X <- mtcars$wt        #variable
alpha <- 0.05         # confidence level = 1-alpha = 0.95
mu <- 3.2             # expectation (arbitrary value)       
pop_var <- 1/(valid.n(X, na.rm=TRUE))*sum((X - mu)^2)  # population variance

# exact CI
IC_min <- pop_var*n/qchisq((1-alpha/2), n)
IC_max <- pop_var*n/qchisq((alpha/2), n)
pop_var
IC_min
IC_max
```
  


# III)  Confidence interval of a proportion
Several methods exists, most often used are:


## II.1) Asymptotic confidence interval 

$$\left[ \overline{X_n} - \frac{1}{2\sqrt{n}}Z_{1-\alpha/2} ; \overline{X_n} + \frac{1}{2\sqrt{n}}Z_{1-\alpha/2} \right] $$
With $\overline{X_n}$ the estimation of the proportion  



```{r}
X <- mtcars$wt        # random variable
alpha <- 0.05         # confidence level = 1-alpha = 0.95
X_n_bar <- 0.3        # estimation of the proportion
n <- (valid.n(X, na.rm=TRUE))
IC_min <- X_n_bar - 1/(2*sqrt(n)*qnorm(1-alpha/2))
IC_max <- X_n_bar + 1/(2*sqrt(n)*qnorm(1-alpha/2))
X_n_bar
IC_min
IC_max
```


## II.2) Agresti-coull Confidence interval

$$\left[ \overline{X_n} - \sqrt{\frac{\overline{X_n}.(1-\overline{X_n})}{n}}Z_{1-\alpha/2} ; \overline{X_n} + \sqrt{\frac{\overline{X_n}.(1-\overline{X_n})}{n}}Z_{1-\alpha/2} \right] $$
With $\overline{X_n}$ the estimation of the proportion

```{r}
X <- mtcars$wt        # random variable
alpha <- 0.05         # confidence level = 1-alpha = 0.95
X_n_bar <- 0.3        # estimation of the proportion
n <- (valid.n(X, na.rm=TRUE))
IC_min <- X_n_bar - sqrt(X_n_bar*(1-X_n_bar)/n)*qnorm(1-alpha/2)
IC_max <- X_n_bar + sqrt(X_n_bar*(1-X_n_bar)/n)*qnorm(1-alpha/2)
X_n_bar
IC_min
IC_max
```
  


# IV)  Confidence interval of the parameter $\lambda$ of an exponential distribution


$$\left[ \frac{1-\frac{Z_{1-\alpha/2}}{\sqrt{n}}}{\overline{X_n}} ; \frac{1+\frac{Z_{1-\alpha/2}}{\sqrt{n}}}{\overline{X_n}}  \right] $$
With $\overline{X_n}$ the empirical mean

```{r}
X <- seq(from = 0, to = 8, by = 0.02)   # (sequence, for simulation) 
X <- dexp(X, rate = 2)#  simulation of a random variable
alpha <- 0.05         # confidence level = 1-alpha = 0.95
lambda_hat <- 1/mean(X)  # estimation of lambda
n <- (valid.n(X, na.rm=TRUE))
IC_min <- (1-qnorm(1-alpha/2)/sqrt(n))/mean(X)
IC_max <- (1+qnorm(1-alpha/2)/sqrt(n))/mean(X)
lambda_hat
IC_min
IC_max
```




