---
title: "Distribution Fitting and Point Estimation"
author: "Benoit Mialet"
date: "02/12/2021"
output:
  pdf_document: default
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Working directory setting:
```{r}
setwd("D:/Formations/DSTI/2021 07 - Advanced stats and ML/assignment")
```


***
Observations of data1.txt dataset are associated to a point process. 
The time arrival between two points should be an exponential random variable with parameter $\lambda$. 
My objective is to find if it is the case with different methods.



To answer the question I will proceed as following:

1. Import data file into a dataframe, and look at the data.
2. Create a new variable *time*, based on the time arrival between two points. The objective will be to determine if the random variable *time* follows  or not an exponential distribution.
3. For the random variable *time*, compute an estimator $\hat{\lambda}$ for the parameter $\lambda$ of a supposed exponential distribution. 2 methods will be used for this purpose:
    + Methods of Moments of orders $k=1$ and $k=2$
    + Method of Maximum Likelihood
4. Determine if the distribution of *time* fits an exponential distribution, with parameter $\hat{\lambda}$. 2 methods will be used for this purpose:
    + Kolmogorov-Smirnov test
    + Visualization of PDF and QQ-Plots, and comparison of the random variable *time* distribution and a theoretical exponential distribution with parameter $\hat{\lambda}$.
5. Conclude 




## I) Data handling and first look to the data


I open *data1.txt* file and store it in a dataframe:
```{r}
data1 <- read.table("data1.txt", header=FALSE, col.names = "point", 
                    sep=";",dec=".", fileEncoding="latin1", check.names=FALSE)

```



I observe some basic information about the data: dimension of the dataframe, type (class) of variables, number of NA values, and first observations
```{r}
str(data1)
paste("Number of NA values : ", sum(is.na(data1)))

head(data1,10)
```
Here, variable *point* is numerical with no missing value.  




## II) Computation of the *time* variable, observation of its metrics and its distribution

To study the the time arrival between points, I create a new dataframe *df_time* from data1 observations, based on the time difference between two successive points. I name this new variable "*time*" :
```{r}
time <- {}
for (i in c(1:(length(data1$point)-1))) {
  time[i] <- data1$point[i+1]-data1$point[i]
}
df_time <- data.frame(time)
head(df_time,10)
```



I observe dimensions and some basic metrics of theses distances (*time*)
```{r}
dim(df_time)
summary(df_time)
```



I plot a histogram of distances to observe the shape of the probability density function of the variable *time* to study:
```{r}
hist(
  df_time$time, 
  breaks = 100, 
  freq=FALSE, 
  main ="Histogram of time arrivals", 
  xlab="Time arrival between points (t)"
  )
```
The distribution of *time* looks like an exponential distribution. In the following, I will first estimate a parameter $\lambda$ before testing the goodness of fit of this distribution with an exponential one.
 
 
 
 
## II) Point estimation: parameter $\lambda$ of an exponential distribution

Let assume that all observations of the dataframe *df_time* $X_1, .., X_n$ are **independent and identically distributed** random variables. Expectation of a random variable $E[X_i]$ with an exponential distribution is by definition:

$$E[X_i]=E[X_1]=\frac{1}{\lambda}$$
With $\lambda \in \mathbb{R}$.

In order to provide some estimators of $\lambda$, I will use several methods in the following.


### II.1) Method of moment with k = 1:
I have:

$$E[X_1^1]=\frac{1}{\lambda}$$

By application of the Law of Large Numbers, an estimator $\hat{\lambda_n}$ for $\lambda$ will be a solution of:

$$g(\hat{\lambda_n})= \frac{1}{\hat{\lambda_n}} = \frac{1}{n}\sum_{i=1}^{n}(X_i) = \overline{X_n}$$
I thus obtain:
$$\hat{\lambda_n} = \frac{1}{\overline{X_n}}$$
I can now compute the value of $\hat{\lambda_n}$:
```{r}
lambda_hat_n <- 1/mean(df_time$time)
lambda_hat_n
```



### II.2) Method of moment with k = 2:

By definition, assuming that $X_1, .., X_n$ are independent and identically distributed random variables, I have by definition:

$$
V[X_1]=E[X_1^2]-E[X_1]^2\\
E[X_1^2]=V[X_1]+E[X_1]^2\\
$$

With $V[X_1]=\frac{1}{\lambda^2}$ and $E[X_1]^2=\frac{1}{\lambda^2}$ (with $\lambda \in \mathbb{R}$).



I thus obtain:

$$E[X_1]^2=\frac{2}{\lambda^2}$$

By application of the Law of Large Numbers, an estimator $\hat{\lambda_{n,2}}$ for $\lambda$ will be a solution of:

$$g(\hat{\lambda_{n,2}})= \frac{2}{\hat{\lambda_{n,2}^2}} = \frac{1}{n}\sum_{i=1}^{n}(X_i^2) = \overline{X_n}$$

I obtain:
$$\hat{\lambda_{n,2}} = \sqrt\frac{2n}{\sum_{i=1}^{n}(X_i^2)}$$


I thus compute $\sum_{i=1}^{n}(X_i^2)$ then $\hat{\lambda_{n,2}}$:
```{r}
sum_Xi_squared <- 0
for (i in c(1:length(df_time$time))) {
  sum_Xi_squared <- sum_Xi_squared + df_time$time[i]**2
}

lambda_hat_n_2 <- sqrt(2*length(df_time$time)/sum_Xi_squared)
lambda_hat_n_2
```



### II.3) Method of Maximum Likelihood:

By definition, probability density function of a continuous random variable $X_i$ which has an exponential distribution is given by:

$$f_{Xi}(x_i)=\lambda.\mathrm{e}^{-\lambda x_i}\mathbb{1}_{[0;+\infty[^{(x_i)}}$$

With $\lambda \in \mathbb{R^{++}}$.


An estimator $\hat{\lambda}$ of $\lambda$ with the method of maximum likelihood would be a solution of the following maximization problem:

$$\hat{\lambda_n} = \arg\max_{\lambda_n} L_{(\lambda_n;x_1,...,x_n)} $$


I assume that all the terms $x_i$ are independent and identically distributed, I can thus write the likelihood function as:

$$
\begin{aligned}
L_{(\lambda;x_1,...,x_n)} &= \prod_{i = 1}^{n} f_{Xi}(x_i;\lambda)\\
&= \prod_{i = 1}^{n}  \left( \lambda\mathrm{e}{-\lambda x_i}\mathbb{1}_{[0;+\infty[^{(x_i)}} \right)\\
&= \lambda^n\mathrm{e}{-\lambda \sum_{i=1}^{n}x_i}\mathbb{1}_{[0;+\infty[^{(min(x_i))}}\\
\end{aligned}
$$

As in this case I am in the frame of an exponential distribution, I have $min(x_i) \ge 0$. I can thus consider only the left part of the equation, without the indicator function:
$$L_{(\lambda;x_1,...,x_n)}=\lambda^n\mathrm{e}{-\lambda \sum_{i=1}^{n}x_i}$$



I can consider a log transformation of the exponential function. I have:

$$
\begin{aligned}
l_{(\lambda;x_1,...,x_n)} &=\ln{(L_{(\lambda;x_1,...,x_n)})}\\
&=\ln{(\lambda^n\mathrm{e}{-\lambda \sum_{i=1}^{n}x_i})}\\
 &= n \ln({\lambda}) - \lambda \sum_{i=1}^{n}x_i\\
\end{aligned}
$$


At a critical point, the derivative of the function will be equal to 0:
$$\frac {d}{d{\lambda_n}} l_{(\lambda_n;x_1,...,x_n)} = 0 $$
I thus have:

$$
\begin{aligned}
\frac {d}{d{\lambda_n}} \left( n \ln({{\lambda_n}}) - {\lambda_n} \sum_{i=1}^{n}x_i \right) &= 0\\
\frac {n}{{\lambda_n}} - \sum_{i=1}^{n}x_i &= 0\\
\end{aligned}
$$



To verify that this solution (critical point) is associated to a maximum, I have to verify that its derivative is $<0$:

$$\frac {d}{d{\lambda_n}} \left( \frac {n}{{\lambda_n}} - \sum_{i=1}^{n}x_i \right) < 0$$

I obtain:
$$ \frac {-n}{{\lambda_n^2}} < 0$$
As $n \ge 0$ and  $\lambda_n^2 \ge 0$, this equation is verified, and I am sure that the previous critical point is associated to a maximum. Thus, an estimator $\hat{\lambda_n}$ of $\lambda$ obtained by the maximum likelihood method is:

$$ \hat{\lambda_n} = \frac {n}{\sum_{i=1}^{n}x_i}$$
The expression can also be written:
$$ \hat{\lambda_n} = \frac {1}{\overline{X_n}}$$
This estimator is the same as the one obtained by method of moments with $k=1$. In the following, I will thus only consider the latter estimator, and will thus compute values for both $\hat{\lambda_n}$ and $\hat{\lambda_{n,2}}$


## IV) Distribution fitting: comparison of time arrivals distribution with an exponential distribution


### IV.1)  Goodness of fit: Kolmogorv-Smirnov test

I compute a Kolmogorv-Smirnov test to compare both distributions, with a Null hypothesis $\mathcal{H}_0$ stating that the observed distribution is not different from an exponential distribution with parameter lambda_hat_n. I will arbitrary set a risk level $\alpha=0.05$.
I will perform one test for each estimator of $\lambda$ I previously computed:
```{r}
ks.test(df_time$time, "pexp", lambda_hat_n)
ks.test(df_time$time, "pexp", lambda_hat_n_2)
```
With p-values 0.1981 and 0.5568 > $\alpha$, I can accept the null hypothesis and assume that time arrivals follows an exponential distribution, for all $\lambda$ estimators previously computed. 





### IV.2)  Histogram and quantiles visualization

In order to visualize and confirm this result, I plot a graph comparing histogram of time arrivals with the probability density functions of exponential distributions with parameters $\hat{\lambda_n}$ (in red) and $\hat{\lambda_{n,2}}$ (in blue):
```{r}
hist(
  df_time$time, breaks = 100, 
  freq=FALSE, 
  main ="Histogram of time arrivals", 
  xlab="Time arrival between points (t)"
  )
curve(dexp(x, rate = lambda_hat_n), from = 0, col = "red", add = TRUE)
curve(dexp(x, rate = lambda_hat_n_2), from = 0, col = "blue", add = TRUE)
```


The curves seems indeed to fit the distribution in both cases. 



I now visualize the Q-Q Plot to compare Quantile values of the *time* variable with quantile values of an exponential distribution:  


```{r}
library(SMPracticals)

# QQ-PLOT
qqexp(df_time$time, main = "Exponential Q-Q Plot",
       plot.it = TRUE, line = TRUE)
```
Again, the distribution of variable *time* seems to really fit an exponential distribution.

## V)  Conclusion

In conclusion, considering result of the Kolmogorv-Smirnov test and by visualization of its histogram and its Q-Q Plot, I can assume that the time arrival between two points is an exponential random variable with parameter $\lambda$. I computed two estimators  $\hat{\lambda_n}$ and $\hat{\lambda_{n,2}}$.

I can thus assume that observations from data1.txt dataset are associated to a Poisson process with parameter $\lambda$.



